**Accuracy and precision of 16S rRNA gene sequence assignments to OTUs is more important than the stability of OTUs**

### 2015-10-02 Schloss_Cluster_PeerJ_2015.Rmd
Work on conclusion section

---
title: "**Accuracy and precision of 16S rRNA gene sequence assignments to OTUs is more important than the stability of OTUs**"
author: Patrick D. Schloss
date: '`r format(Sys.Date(), format="%B %d, %Y")`'
bibliography: references.bib
output:
  pdf_document:
    includes:
      in_header: header.tex
csl: peerj.csl
fontsize: 11pt
geometry: margin=1.0in
---

@@

inline_hook <- function(x){
  if(is.numeric(x)){
    paste(format(x,digits=2))
  }
}
knitr::knit_hooks$set(big.sep=',')
```
\newpage

@@

* Reference-based clustering has many problems
* Open-reference clustering issues
* Implemented in numerous pipelines for large projects

* Many difficulties in clustering have to do with underlying data quality.

>>>>

---
title: "**Accuracy and precision of 16S rRNA gene sequence assignments to OTUs is more important than the stability of OTUs**"
author: Patrick D. Schloss* and Sarah L. Westcott
date: '`r format(Sys.Date(), format="%B %d, %Y")`'
bibliography: references.bib
output:
  pdf_document:
    includes:
      in_header: header.tex
csl: peerj.csl
fontsize: 11pt
geometry: margin=1.0in
---

@@

inline_hook <- function(x){
  if(is.numeric(x)){
    paste(format(x,big.mark=','))
  }
}
knitr::knit_hooks$set(inline=inline_hook)
```

* To whom correspondence should be addressed.
pschloss@umich.edu

Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI

@@

For the first time, we assessed the ability of reference-based clustering algorithms to capture the actual similarity of the sequences in a dataset. Several studies have lauded both the open and closed-reference approaches for generating reproducible clusterings, yet we showed that both reference-based approaches did a poor job of representing the similarity between the sequences compared to the *de novo* approaches. So, although the clusterings are reproducible and stable across a range of library sizes, the clusterings are a poor representation of the data. We also observed that the clusterings were not actually reproducible when the order of the reference sequences was randomize. When USEARCH was used, the actual number of sequences that mapped to the reference changed depending on the order of the reference. Perhaps most alarming is that the default order of the database provided the worst sensitivity of any of the randomizations we attempted. Even when we used VSEARCH to perform closed-reference clustering and were able to obtain a consistent clusterings, we observed that the labels on the OTUs differed between randomizations. Because the OTU labels are frequently used to identify representative sequences for those OTUs, variation in labels, often representing different taxonomic groups, will have a detrimental effect on the interpretation of downstream analyses.

Because the open-reference algorithm is a hybrid of the closed-reference and DGC algorithms, it is also negatively affected by the problems with using USEARCH for both algorithms. An added problem with the open-reference algorithm is that the two phases of the algorithm employ different thresholds. In the closed-reference step, sequences must be within a threshold of a reference to be in the same OTU. This means that two sequences that are 97% similar to a reference and are joined into the same OTU, may only be 94% similar to each other. In the DGC step, the goal is to approximate the AL algorithm which requires that, on average, the sequences within an OTU are at least 97% similar to each other. The end result of the open-reference approach is that sequences that are similar to previously observed sequences are clustered with one threshold while those that are not similar to previously observed sequences are clustered with a different threshold.

As the throughput of sequencing technologies have improved, algorithm development to keep pace. *De novo* clustering algorithms are considerably slower and more computationally intensive than reference-based algorithms and the greedy *de novo* algorithms are faster than the hierarchical. In our experience (REFS), the most significant detriment to execution speed of the *de novo* algorithms has been in adequate removal of sequencing error. As the rate of sequencing error increases so do the number of unique sequences that must be clustered. The speed of the *de novo* algorithms scales approximately quadratically, so that doubling the number of sequences results in a four-fold increase in the time required to execute the algorithm. The rapid expansion in sequencing throughput has been likened to the Red Queen in Lewis Carroll's, *Through the Looking-Glass* who must run in place to keep up to her changing surroundings (REF). Microbial ecologists must continue to refine clustering algorithms to better handle the size of the datasets, but they must also take steps to improve the quality of the underlying data. Ultimately, objective standards must be applied to assess the quality of the data and the quality of OTU clustering.
